A performance comparison of Dask and Apache Spark for data-intensive neuroimaging pipelines

### 2019-07-29 latex/paper.tex

In the past few years, neuroimaging has entered the Big Data era due to the
joint increase in image resolution, data sharing, and study sizes. However,
no particular Big Data framework has emerged in this field, and several
alternatives remain available. We compare two popular Big Data frameworks
with Python APIs, Apache Spark and Dask, for their runtime performance in
processing neuroimaging pipelines. Our evaluation uses two synthetic
pipelines processing the 81~GB Big Brain image, and a real pipeline
processing anatomical data from more than 1,000 subjects. We benchmark these pipelines using
 various combinations of task durations, data sizes, and numbers of workers,
deployed on a 64-node compute cluster in Compute Canada's Arbutus
cloud. We evaluate PySpark's RDD API against Dask's Bag, Delayed and Futures.
Results show that despite slight differences between Spark and Dask, both
frameworks perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all
cases, the major limiting factor was data transfer. While either framework
is suitable for neuroimaging pipelines, more effort needs to be placed in
reducing data transfer time.
\end{abstract}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data frameworks, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

%Similarities
Spark and Dask both offer in-memory computing, data locality, and lazy evaluation,
which is common for Big Data frameworks. Both their schedulers operate dynamically,
which is good when task runtimes are not known ahead of time~\cite{Dask:15}. They
also provide rich, high-level programming APIs, and support a variety of
infrastructure schedulers, such as Mesos~\cite{hindman2011mesos}, YARN~\cite{vavilapalli2013apache}, or HPC clusters (Dask only). Over
these similarities, the frameworks have differences.

% Differences
First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss. Our study focuses on
performance, although we recognize that other factors will also impact the
choice of a framework in practice.

% Related work
Spark and Dask were both included in the evaluation reported
in~\cite{Mehta:17}, where a neuroimaging application
processed approximately \SI{100}{\giga\byte} of data. In this work, Dask was
reported to have a slight performance advantage over Spark. Overall, Dask's
end-to-end time (makespan) was measured to be up to 14\% faster than Spark,
due to ``more efficient pipelining'' %(Fig 10) 
and serialization time to
Python. % (Fig. 12-a)
Dask, however, was reported to have a larger startup
time than Spark. % (Fig. 12-c)
The analysis remained at a quite high level
though, leaving most of the observed performance difference unexplained. In
comparison, our study will provide a detailed analysis of performance
differences and similarities between Spark and Dask for neuroimaging data processing.

@@

Figure~\ref{fig:inc_tt_worker} shows the total execution time of the
Incrementation application, broken down into data transfer (read and
write), compute (sleep), and overhead time. As expected, the computing time
stays similar when the number of workers increases. However, the data
transfer time and overhead increase substantially with the number of
workers.

% \hspace{1cm}\(\hat{y}=2336.9499\cdot{x} + 325.00574\) (Spark)

% \hspace{1cm}\(\hat{y}=2650.13505\cdot{x} + 1585.37218\) (Bag)

% \hspace{1cm}\(\hat{y}=3250.53406\cdot{x} + 1769.50351\) (Delayed)

% \hspace{1cm}\(\hat{y}=3570.39758\cdot{x} + 1062.23031\) (Futures)

@@

In Figure~\ref{fig:histo_tt_block} the total time for each function is shown.
Note that at 30 blocks the workers' resources are not fully used since the
workers can process up to 64 tasks in parallel. Moreover, the overhead for
30 blocks is erroneous due to a lower amount of block than available threads.
This leads to thread based engines to have higher overhead. Given that, 
Spark has a much lower compute time which makes it significantly faster
than Dask engines. Overall, for all engines, lower block size results in lower
IO time but increases the overhead time. Again, this is due to large overhead
desynchronizing task thus reducing the NFS bottleneck.

@@

Figure~\ref{fig:histo_np_tt_block} shows the total time spent in each function.
Once more, since Delayed is thread-based its overhead at 30 blocks is seems much
higher that it really is. Considering the overhead error and that at 30 blocks
the workers only use half their core, the total time is substantially the same
for all block size. This is due to lower block sizes having a lower IO time,
however, it balances out by the resulting increase in overhead.

@@

In Figure~\ref{fig:histo_np_gantt}, read tasks are initially similar for all
engines. However, when computation starts the subsequent read tasks are much
shorter for Delayed. This is due the desynchronization of the task reducing the
IO bottleneck.

@@

In Figure~\ref{fig:bids_tt_worker}, the total execution time of each function is
shown. Each engine has a similar total time independently of its number of workers.
Consistently with previous observations, the overhead increases substantially with the 
number of workers.

% The overhead increases
% proportionally with the number of workers (see regression lines below). This is due
% to more communication between the scheduler and workers as well as intra-worker and
% inter-worker communication.

% \hspace{1cm}\(\hat{y}=223.53586\cdot{x} + 91.616964\) (Spark)

% \hspace{1cm}\(\hat{y}=182.82659\cdot{x} - 11.43834\) (Bag)

% \hspace{1cm}\(\hat{y}=172.54183\cdot + 400.69433\) (Delayed)

% \hspace{1cm}\(\hat{y}=132.41481\cdot{x} + 90.32813\) (Futures)

@@

\subsection{Effect of Network File System}
We used an NFS as it is a common approach to store a large amount of data. However,
it limits the available bandwidth since all workers interact with one disk. This
causes data-intensive applications to take a significant amount of time for IO. As we
expect, this effect is amplified when too many tasks simultaneously try to operate
with the NFS at the same time as
Fig.~\ref{fig:inc_gantt},\ref{fig:histo_gantt},\ref{fig:histo_np_gantt},\ref{fig:bids_gantt}
shows. A parallel file system, such as Lustre~\cite{lustre}, would likely improve
IO times of parallel applications simultaneously writing to shared storage.

@@

Data transfer accounts for the majority of the execution time for data-intensive
applications (see
Fig.~\ref{fig:inc_gantt},\ref{fig:histo_spark_gantt},\ref{fig:histo_np_gantt}). We
see that when a low number of IO tasks are overlapping, this diminishes the bandwidth
limitation from the NFS. We want to explore solutions to desynchronize IO tasks in
the aim to reduce the makespan of data-intensive applications.

@@

\subsection{Effect of tasks' size}
We measured the effect of various block sizes for constant BigBrain image for
Incrementation, Histogram, and Histogram (NumPy). We observe that larger block size
results in more IO compensated by lower overhead time. On the other hand, smaller
block size results in higher overhead time compensated by reduced IO time.

@@

Varying the block size is equivalent to varying the size of the partitions. We
observed no difference when varying the task size while using an NFS. However, if
one uses a parallel shared file system, we suppose it would be preferable to use a
larger block size as the IO time should be reduced substantially. It is worth noting
that the Dask developers suggest, in
\href{https://docs.dask.org/en/latest/best-practices.html}{Dask's best practices}, to
use a partition size such that workers have 2-3 times as many chunks available to
work on. More experiments would have to be performed to gain insight on an optimal
number of partitions to use for a given set of data-intensive applications.

>>>>

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines has emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} Big Brain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on a 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both engines perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either engine is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data engines, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

%Similarities
Spark and Dask both offer in-memory computing, data locality, and lazy evaluation,
which is common for Big Data engines. Both their schedulers operate dynamically,
which is good when task runtimes are not known ahead of time~\cite{Dask:15}. They
also provide rich, high-level programming APIs, and support a variety of
infrastructure schedulers, such as Mesos~\cite{hindman2011mesos}, YARN~\cite{vavilapalli2013apache}, or HPC clusters (Dask only). Over
these similarities, the engines have differences.

First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss. Our study focuses on
performance, although we recognize that other factors will also impact the
choice of an engine in practice.

Spark and Dask were both included in the evaluation reported
in~\cite{Mehta:17}, where a neuroimaging application
processed approximately \SI{100}{\giga\byte} of data. In this work, Dask was
reported to have a slight performance advantage over Spark. Overall, Dask's
end-to-end time (makespan) was measured to be up to 14\% faster than Spark,
due to ``more efficient pipelining'' %(Fig 10) 
and serialization time to
Python. % (Fig. 12-a)
Dask, however, was reported to have a larger startup
time than Spark. % (Fig. 12-c)
The analysis remained at a quite high level
though, leaving most of the observed performance difference unexplained. In
comparison, our study will provide a detailed analysis of performance
differences and similarities between Spark and Dask for neuroimaging data processing.

@@

Figure~\ref{fig:inc_tt_worker} shows the total execution time of the
Incrementation application, broken down into data transfer (read and
write), compute (sleep), and overhead time. As expected, the computing time
stays similar when the number of workers increases. However, the data
transfer time and overhead increase proportionally to the number of
workers with regression slopes: Spark (\SI{2337}{\second/task}), Bag
(\SI{2650}{\second/task}), Delayed (\SI{3251}{\second/task}), Futures
(\SI{3570}{\second/task}).

@@

In Figure~\ref{fig:histo_tt_block} the total time for each function is shown.
Note that at 30 blocks the workers' resources are not fully used since the
workers can process up to 64 tasks in parallel. Moreover, the overhead for
30 blocks is erroneous due to a lower amount of block than available threads.
This leads to thread based engines to have higher overhead. Given that, 
Spark has a much lower compute time which makes it significantly faster
than Dask engines. Overall, for all engines, lower block size results in lower
data transfer time but increases the overhead time. Again, this is due to large overhead
desynchronizing task thus reducing the NFS bottleneck.

@@

Figure~\ref{fig:histo_np_tt_block} shows the total time spent in each function.
Once more, since Delayed is thread-based its overhead at 30 blocks is seems much
higher that it really is. Considering the overhead error and that at 30 blocks
the workers only use half their core, the total time is substantially the same
for all block size. This is due to lower block sizes having a lower data transfer time,
however, it balances out by the resulting increase in overhead.

@@

In Figure~\ref{fig:histo_np_gantt}, read tasks are initially similar for all
engines. However, when computation starts the subsequent read tasks are much
shorter for Delayed. This is due the desynchronization of the task reducing the
data transfer bottleneck.

@@

In Figure~\ref{fig:bids_tt_worker}, the total execution time of each function is
shown. Each engine has a similar total time independently of its number of workers.
This is due to the, almost, linear scaling of the engines. The overhead increases
proportionally to the number of workers with regression slopes: Spark
(\SI{224}{\second/task}), Bag (\SI{183}{\second/task}), Delayed
(\SI{173}{\second/task}), Futures (\SI{132}{\second/task}). This is due to more
communication between the scheduler and workers as well as intra-worker and
inter-worker communication.

@@

\subsection{Effect of Network File System}
We used an NFS as it is a common approach to store a large amount of data. However,
it limits the available bandwidth since all workers interact with one disk. This
causes data-intensive applications to take a significant amount of time for data transfer. As we
expect, this effect is amplified when too many tasks simultaneously try to operate
with the NFS at the same time as
Fig.~\ref{fig:inc_gantt},\ref{fig:histo_gantt},\ref{fig:histo_np_gantt},\ref{fig:bids_gantt}
shows. A parallel file system, such as Lustre~\cite{lustre}, would likely improve
data transfer times of parallel applications simultaneously writing to shared storage.

@@

Data transfer accounts for the majority of the execution time for data-intensive
applications (see
Fig.~\ref{fig:inc_gantt},\ref{fig:histo_spark_gantt},\ref{fig:histo_np_gantt}). We
see that when a low number of data transfer tasks are overlapping, this diminishes the bandwidth
limitation from the NFS. We want to explore solutions to desynchronize data transfer tasks in
the aim to reduce the makespan of data-intensive applications.

@@

\subsection{Effect of tasks' size}
We measured the effect of various block sizes for constant BigBrain image for
Incrementation, Histogram, and Histogram (NumPy). We observe that larger block size
results in more data transfer compensated by lower overhead time and vice-versa.

@@

Varying the block size or the size of the partitions is equivalent. No difference is
observed when varying the task size while using an NFS. However, if one uses a
parallel shared file system, it would be preferable to use a larger block
size as the data transfer time should be reduced substantially. It is worth noting that the Dask
developers suggest, in
\href{https://docs.dask.org/en/latest/best-practices.html}{Dask's best practices}, to
use a partition size such that workers have 2-3 times as many chunks available to
work on. More experiments would have to be performed to gain insight on an optimal
number of partitions to use for a given set of data-intensive applications.
